---
title: "Anomaly_EDA"
author: "Andrew Jossi"
date: "2024-10-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(glmnet)
library(lubridate)
library(stringr)
library(readxl)
```

## Model

```{r, echo=FALSE}
modeldata <- read_csv("modeldata.csv")
#View(modeldata)

#add a datetime column
modeldata$datetime <- as.POSIXct(
  paste(modeldata$year.y, modeldata$month.y, modeldata$day.y, modeldata$hour.y, modeldata$minute.y),
  format = "%Y %m %d %H %M",
  tz = "UTC"  
)

```

Created a subset of the modeldata for June 2018 as there were 57 anomalies in that month
```{r, echo=FALSE}
june_2018_data <- subset(modeldata, month.y == 6 & year.y == 2018)

#head(june_2018_data)
```


## Small Irradiances Graphed over time with Anomalies
```{r, echo=FALSE}
ggplot(june_2018_data, aes(x = datetime, y = irr_256)) +
  geom_line(color = "blue") +                           # Line for irr_256 over time
  geom_point(data = subset(june_2018_data, binary == 1), 
             aes(x = datetime, y = irr_256), 
             color = "red", size = 2) +                 # Red points for anomalies
  labs(title = "irr_256 over Time with Anomalies Highlighted",
       x = "Time",
       y = "irr_256") +
  theme_minimal()
```
No apparent relationship between irr_256 and anomalies. It appears as though two solar events occur in this month (solar flare, cme,...). NOAA's alert system sent out alerts for a Geomagnetic storm with K-index of 5 in the begining of the month and an Electron 2MeV Integral Flux exceeded 1000pfu alert in the latter part of the month. 


## Large Irradiances Graphed over time with Anomalies
```{r, echo=FALSE}
ggplot(june_2018_data, aes(x = datetime, y = irr_1175)) +
  geom_line(color = "blue") +                           # Line for irr_1175 over time
  geom_point(data = subset(june_2018_data, binary == 1), 
             aes(x = datetime, y = irr_1175), 
             color = "red", size = 2) +                 # Red points for anomalies
  labs(title = "irr_1175 over Time with Anomalies Highlighted",
       x = "Time",
       y = "irr_1175") +
  theme_minimal()
```
Anomalies appear to occur when many spikes happen in a short window of time, about 7 days.

```{r, echo=FALSE}
ggplot(june_2018_data, aes(x = datetime, y = irr_1216)) +
  geom_line(color = "blue") +                           # Line for irr_1216 over time
  geom_point(data = subset(june_2018_data, binary == 1), 
             aes(x = datetime, y = irr_1216), 
             color = "red", size = 2) +                 # Red points for anomalies
  labs(title = "irr_1216 over Time with Anomalies Highlighted",
       x = "Time",
       y = "irr_1216") +
  theme_minimal()
```
Higher irradiance spikes seems to be more related to anomalies, as anomalies occur at nearly every spike. 


```{r, echo=FALSE}
ggplot(june_2018_data, aes(x = datetime, y = irr_1335)) +
  geom_line(color = "blue") +                           # Line for irr_1335 over time
  geom_point(data = subset(june_2018_data, binary == 1), 
             aes(x = datetime, y = irr_1335), 
             color = "red", size = 2) +                 # Red points for anomalies
  labs(title = "irr_1335 over Time with Anomalies Highlighted",
       x = "Time",
       y = "irr_1335") +
  theme_minimal()
```


```{r, echo=FALSE}
ggplot(june_2018_data, aes(x = datetime, y = irr_1405)) +
  geom_line(color = "blue") +                           # Line for irr_1405 over time
  geom_point(data = subset(june_2018_data, binary == 1), 
             aes(x = datetime, y = irr_1405), 
             color = "red", size = 2) +                 # Red points for anomalies
  labs(title = "irr_1405 over Time with Anomalies Highlighted",
       x = "Time",
       y = "irr_1405") +
  theme_minimal()
```

## MGII graphs
```{r, echo=FALSE}
ggplot(june_2018_data, aes(x = datetime, y = MgII_EXIS)) +
  geom_line(color = "blue") +                           # Line for MgII_EXIS  over time
  geom_point(data = subset(june_2018_data, binary == 1), 
             aes(x = datetime, y = MgII_EXIS), 
             color = "red", size = 2) +                 # Red points for anomalies
  labs(title = "MgII_EXIS over Time with Anomalies Highlighted",
       x = "Time",
       y = "MgII_EXIS") +
  theme_minimal()
```

```{r, echo=FALSE}
ggplot(june_2018_data, aes(x = datetime, y = MgII_standard)) +
  geom_line(color = "blue") +                           # Line for MgII_standard  over time
  geom_point(data = subset(june_2018_data, binary == 1), 
             aes(x = datetime, y = MgII_standard), 
             color = "red", size = 2) +                 # Red points for anomalies
  labs(title = "MgII_standard over Time with Anomalies Highlighted",
       x = "Time",
       y = "MgII_standard") +
  theme_minimal()

```
MgII seems to have a lagged correlation with anomalies. They seem to occur during the decrease of MgII after a large spike.


## May 2018 data

Feel free to change the month and year assignment to explore other time periods
```{r, echo=FALSE}
may_2018_data <- subset(modeldata, month.y == 5 & year.y == 2018)

#head(may_2018_data)
```


```{r, echo=FALSE}
ggplot(may_2018_data, aes(x = datetime, y = irr_1175)) +
  geom_line(color = "blue") +                           # Line for irr_1175  over time
  geom_point(data = subset(may_2018_data, binary == 1), 
             aes(x = datetime, y = irr_1175), 
             color = "red", size = 2) +                 # Red points for anomalies
  labs(title = "irr_1175 over Time with Anomalies Highlighted",
       x = "Time",
       y = "irr_1175") +
  theme_minimal()
```



```{r, echo=FALSE}
ggplot(may_2018_data, aes(x = datetime, y = irr_1405)) +
  geom_line(color = "blue") +
  geom_point(data = subset(may_2018_data, binary == 1), 
             aes(x = datetime, y = irr_1405), 
             color = "red", size = 2) +
  labs(title = "irr_1405 over Time with Anomalies Highlighted",
       x = "Datetime",
       y = "irr_1405") +
  theme_minimal()
```

# Whole data set (4 year) patterns
```{r, echo=FALSE, eval=FALSE}
colnames(modeldata)
```
Change this graph to any of the above column names to explore relationships. All `irr_###` columns relate to Extreme Ultraviolet Light (EUV) data and xrsb relates to x-ray data. 

```{r, echo=FALSE}
ggplot(modeldata, aes(x = datetime, y = irr_1405)) +
  geom_line(color = "blue") +                           # Line for irr_1405 over time
  geom_point(data = subset(modeldata, binary == 1), 
             aes(x = datetime, y = irr_1405), 
             color = "red", size = 2) +                 # Red points for anomalies
  labs(title = "irr_1405 over Time with Anomalies Highlighted",
       x = "Time",
       y = "irr_1405") +
  theme_minimal()
```
Wave-like patterns for irradiances. 

```{r, echo=FALSE}
# Assuming modeldata has columns: datetime, irr_1405, irr_256, irr_284, etc.
ggplot(modeldata, aes(x = datetime)) +
  geom_line(aes(y = irr_1405, color = "irr_256")) +    # Line for irr_1405 in blue
  geom_line(aes(y = irr_256, color = "irr_1175")) +      # Line for irr_256 in a different color
  geom_line(aes(y = irr_284, color = "irr_1216")) +      # Line for irr_284 in another color
  geom_line(aes(y = irr_304, color = "irr_1405")) +      # Line for irr_304 in another color
  labs(title = "Irradiance Levels over Time",
       x = "Time",
       y = "Irradiance") +
  scale_color_manual(values = c("irr_256" = "blue", "irr_1175" = "green", "irr_1216" = "purple", "irr_1405" = "orange")) +
  theme_minimal() +
  theme(legend.title = element_blank())

```


```{r, echo=FALSE}
# Assuming modeldata has columns for all specified irradiances and fluxes
ggplot(modeldata, aes(x = datetime)) +
  geom_line(aes(y = irr_1175, color = "irr_1175")) +
  geom_line(aes(y = irr_1405, color = "irr_1405")) +
  labs(title = "Irradiance over Time",
       x = "Time",
       y = "Value") +
  scale_color_manual(values = c(
                                "irr_1175" = "red",
                                "irr_1405" = "blue")) +
  theme_minimal() +
  theme(legend.title = element_blank())
```


## Shorter Period  Patterns (Set duration)

```{r, echo=FALSE}
# Define the start and end date for the 3-month period
start_date <- as.POSIXct("2019-01-01")  # Change to your desired start date
end_date <- as.POSIXct("2019-03-28")    # Change to your desired end date

# Filter the data for the specific 3-month period
modeldata_filtered <- modeldata %>%
  filter(datetime >= start_date & datetime <= end_date)

# Create the plot
ggplot(modeldata_filtered, aes(x = datetime, y = irr_1405)) +
  geom_line(color = "blue") +                           # Line for irr_1216 over time
  geom_point(data = subset(modeldata_filtered, binary == 1), 
             aes(x = datetime, y = irr_1405), 
             color = "red", size = 2) +                 # Red points for anomalies
  labs(title = "irr_1405 over Time with Anomalies Highlighted",
       x = "Time",
       y = "irr_1405") +
  theme_minimal()

```

# Model

```{r}
#split data
set.seed(303) 
sample_index <- sample(seq_len(nrow(modeldata)), size = 0.8 * nrow(modeldata))
train_data <- modeldata[sample_index, ]
test_data <- modeldata[-sample_index, ]


#add sin and cos variables in training data
train_data <- train_data %>%
  mutate(day_of_year = yday(datetime))

train_data <- train_data %>%
  mutate(
    irr_cos = cos(2 * pi * day_of_year / 365)
  )

#add sin and cos in test data
test_data <- test_data %>%
  mutate(day_of_year = yday(datetime))

test_data <- test_data %>%
  mutate(
    irr_cos = cos(2 * pi * day_of_year / 365)
  )


#model_log_cos <- glm(binary ~ xrsb_flux_electrons + xrsa_flux_electrons + irr_cos,data = train_data, family = binomial)

#model_log_eda <- glm(binary ~ xrsb_flux_electrons + xrsa_flux_electrons + xrsa_flag + xrsa_flux_observed + xrsb_flag + irr_cos, data = train_data, family = binomial)

model_log_eda <- glm(binary ~ xrsb_flux_electrons + xrsa_flux_electrons + xrsb_flux_observed + xrsb_flux + irr_cos, data = train_data, family = binomial)

summary(model_log_eda)
```

```{r, echo=FALSE}
# Predict probabilities on the test set
prob_logistic <- predict(model_log_eda, test_data, type = "response")
predictions_logistic <- ifelse(prob_logistic > 0.5, 1, 0)

# Calculate accuracy
accuracy_logistic <- mean(predictions_logistic == test_data$binary)
print(paste("Logistic Regression Accuracy:", accuracy_logistic))

# ROC curve and AUC
library(pROC)
roc_logistic <- roc(test_data$binary, prob_logistic)
auc_logistic <- auc(roc_logistic)
print(paste("Logistic Regression AUC:", auc_logistic))

```


```{r, echo=FALSE}

ggplot(modeldata, aes(x = datetime)) +
  geom_line(aes(y = irr_1405), color = "blue") +  # Set line color to blue
  theme_minimal() +
  labs(title = "irr_1405 over Time",
       x = "Time",
       y = "irr_1405") +
  theme(legend.title = element_blank())

```

```{r}

library(caret)
library(pROC)
library(dplyr)

# Ensure 'binary' column is treated as a factor with two levels
modeldata$binary <- factor(modeldata$binary, levels = c(0, 1), labels = c("No", "Yes"))

# Add cos variable to the entire dataset
modeldata <- modeldata %>%
  mutate(day_of_year = yday(datetime)) %>%
  mutate(irr_cos = cos(2 * pi * day_of_year / 365))

# Set up cross-validation with 5 folds
set.seed(303)
control <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = twoClassSummary)

# Define the formula for the logistic regression model
formula_eda <- binary ~ xrsb_flux_electrons + xrsa_flux_electrons + xrsb_flux_observed + xrsb_flux + irr_cos

# Train the logistic regression model using cross-validation
model_log_cv <- caret::train(
  formula_eda,
  data = modeldata,
  method = "glm",
  family = binomial,
  trControl = control,
  metric = "ROC"
)

# Print cross-validation results
print(model_log_cv)

# Predict probabilities on the full dataset for ROC analysis
prob_log_cv <- predict(model_log_cv, modeldata, type = "prob")[, 2]  # Probability of "Yes"
predictions_log_cv <- ifelse(prob_log_cv > 0.5, "Yes", "No")

# Calculate accuracy on the full dataset (for comparison)
accuracy_log_cv <- mean(predictions_log_cv == modeldata$binary)
print(paste("Cross-Validated Logistic Regression Accuracy:", accuracy_log_cv))

# Calculate ROC curve and AUC
roc_log_cv <- roc(modeldata$binary, prob_log_cv, levels = c("No", "Yes"), direction = "<")
auc_log_cv <- auc(roc_log_cv)
print(paste("Cross-Validated Logistic Regression AUC:", round(auc_log_cv, 4)))

# Plot the ROC curve
plot(roc_log_cv, col = "blue", main = "ROC Curve (Logistic Regression with CV)")

# Plot the ROC curve with reversed x-axis
plot(
  roc_log_cv,
  col = "blue",
  main = "ROC Curve (Logistic Regression with CV)",
  xlim = c(1, 0),  # Reverse x-axis limits
  xlab = "Specificity (1 - False Positive Rate)",
  ylab = "Sensitivity (True Positive Rate)"
)
```

```{r}
library(ROCR)

# Use predictions and labels to calculate the ROC
prob_log_cv <- predict(model_log_cv, modeldata, type = "prob")[, 2]  # Probability of "Yes"
pred <- prediction(prob_log_cv, modeldata$binary)

# Calculate performance
perf <- performance(pred, "tpr", "fpr")

# Plot ROC curve with colorization and reversed x-axis
plot(
  perf, 
  colorize = TRUE,
  main = "ROC Curve (Logistic Regression with CV)",
  xlim = c(0, 1),  # Reverse x-axis
  asp = 0.5
)


```




